---
title: "Reproducible Research: Peer Assessment 1"
author: "Minerva Goree"
date: "Saturday, August 09, 2014"
output: html_document
---
### Global Options for this report
```{r setoptions, echo=TRUE}
        require('knitr')
        opts_chunk$set(echo = TRUE, results = "asis")
        options(digits=2)
```

### Loading and preprocessing the data
```{r loadingData}
##unzip("./activity.zip")
activity <- read.csv("activity.csv", header=TRUE)
#Change dates from factor to date format.
activity$date <- as.character(as.factor(activity$date))
activity$date <- as.Date(activity$date)
```

### What is mean total number of steps taken per day?
```{r meanSteps}
totalSteps <- with(activity, aggregate(steps, by=list(date), FUN=sum, na.rm=TRUE))
names(totalSteps)<- c("date","steps")
stepsHist <- hist(totalSteps$steps,main="Total Daily Steps Histogram",xlab="Total Daily Steps", col="blue",ylim=c(0,20),breaks=10)

stepsMean <-mean(totalSteps$steps)
stepsMedian <- median(totalSteps$steps)
```


- Daily Steps Mean: `r stepsMean`
- Daily Steps Median: `r stepsMedian`

### What is the average daily activity pattern?
```{r dailypattern}
activityData <- activity
activityData$time <- sprintf("%04d:00",activityData$interval)
activityData$time <- gsub("(\\d\\d)(\\d\\d)","\\1:\\2",activityData$time)
activityData$dateTime <- as.POSIXct(paste(activityData$date,activityData$time),format="%Y-%m-%d %H:%M:%S")
activityData$time <- as.POSIXct(activityData$time,format="%H:%M:%S")
interval <- with(activityData, aggregate(steps, by=list(time), FUN=mean, na.rm=TRUE))
names(interval)<- c('time','steps')
plot(interval$time,interval$steps,type="l",col="blue", main="Average Daily Steps per 5 Minute Interval", xlab="Daily Interval by Hour",ylab="Steps")
maxSteps <- which.max(interval[,2])
maxSteps <- interval[maxSteps,]
maxSteps$steps <- round(maxSteps$steps,digits=0)

```
The maximum steps are found on interval (time) **`r maxSteps$time `** at about **`r maxSteps$steps`** steps.

### Imputing missing values
Calculate how many rows have missing values (NAs).

```{r missingValues}
notComplete <- activity[!complete.cases(activity),]
notComplete2 <- nrow(notComplete)
activityRows <- nrow(activity)
percentIncomplete <- notComplete2/activityRows*100
```
The number of incomplete rows in our data set is **`r notComplete2 ` **, which is **`r percentIncomplete` **% of our total data set.

To take care of the missing values I will use the mean value of each interval to fill in the missing NAs.  
```{r fillNA}
notCompleteData <- activity[!complete.cases(activity),]

```

### Are there differences in activity patterns between weekdays and weekends?
